---
title: "Early Childhood Education Access"
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(scales)
  library(broom)
  library(knitr)
  library(sf)
  library(councildown)
})

setDTthreads(1L)

dt <- fread("data/lm_model_data.csv")
dt[, puma := sprintf("%04d", as.integer(puma))]
dt[
  ,
  `:=`(
    occupied_seats = total_seats - total_vacancy,
    total_seats_per_child = total_seats / pre_sch_age,
    occupied_seats_per_child = (total_seats - total_vacancy) / pre_sch_age,
    pct_hispanic10 = pct_hispanic * 10,
    pct_black10 = pct_black_nh * 10,
    pct_asian10 = pct_asian_nh * 10
  )
]

race_categories <- c(
  "White",
  "Black",
  "Hispanic",
  "Asian",
  "Other / 2+"
)

dt[
  ,
  plurality_group := race_categories[
    max.col(
      cbind(
        pct_white_nh,
        pct_black_nh,
        pct_hispanic,
        pct_asian_nh,
        pct_other_2plus_combined
      ),
      ties.method = "first"
    )
  ]
]

total_model <- lm(
  log(total_seats_per_child) ~
    pct_hispanic10 +
    pct_black10 +
    pct_asian10 +
    median_inc_10k,
  data = dt
)

occupied_model <- lm(
  log(occupied_seats_per_child) ~
    pct_hispanic10 +
    pct_black10 +
    pct_asian10 +
    median_inc_10k,
  data = dt
)

fmt_p <- function(p) {
  fcase(
    is.na(p), NA_character_,
    p < 0.001, "<0.001",
    default = sprintf("%.3f", p)
  )
}

prep_results <- function(model, outcome) {
  out <- as.data.table(tidy(model, conf.int = TRUE))
  out[
    ,
    `:=`(
      outcome = outcome,
      percent_change = 100 * (exp(estimate) - 1),
      percent_change_low = 100 * (exp(conf.low) - 1),
      percent_change_high = 100 * (exp(conf.high) - 1),
      term_label = fcase(
        term == "pct_hispanic10", "Hispanic",
        term == "pct_black10", "Black",
        term == "pct_asian10", "Asian",
        term == "median_inc_10k", "Median household income",
        default = term
      )
    )
  ]
  out[term != "(Intercept)"]
}

model_results <- rbindlist(list(
  prep_results(total_model, "Total seats per child"),
  prep_results(occupied_model, "Occupied seats per child")
))

race_results <- model_results[term %in% c("pct_hispanic10", "pct_black10", "pct_asian10")]
occupied_race_results <- race_results[outcome == "Occupied seats per child"]
largest_adjusted <- occupied_race_results[which.max(percent_change)]

snapshot <- data.table(
  Metric = c(
    "PUMAs",
    "Preschool-age children",
    "Total ECE seats",
    "Occupied ECE seats",
    "Mean total seats per child",
    "Mean occupied seats per child",
    "Median household income"
  ),
  Value = c(
    comma(nrow(dt)),
    comma(sum(dt$pre_sch_age, na.rm = TRUE)),
    comma(sum(dt$total_seats, na.rm = TRUE)),
    comma(sum(dt$occupied_seats, na.rm = TRUE)),
    number(mean(dt$total_seats_per_child, na.rm = TRUE), accuracy = 0.01),
    number(mean(dt$occupied_seats_per_child, na.rm = TRUE), accuracy = 0.01),
    dollar(median(dt$median_household_income, na.rm = TRUE), accuracy = 1)
  )
)

composition_summary <- dt[
  ,
  .(
    pumas = .N,
    mean_total_seats_per_child = mean(total_seats_per_child, na.rm = TRUE),
    mean_occupied_seats_per_child = mean(occupied_seats_per_child, na.rm = TRUE)
  ),
  by = plurality_group
]
composition_summary <- merge(
  data.table(plurality_group = race_categories),
  composition_summary,
  by = "plurality_group",
  all.x = TRUE,
  sort = FALSE
)
composition_summary[is.na(pumas), pumas := 0L]
composition_summary <- composition_summary[pumas > 0]

pal <- councildown::pal_nycc("mixed")
warm_pal <- councildown::pal_nycc("warm")

coef_plot <- ggplot(
  race_results,
  aes(
    x = percent_change,
    y = reorder(term_label, percent_change),
    color = outcome
  )
) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_errorbar(
    aes(xmin = percent_change_low, xmax = percent_change_high),
    height = 0.16,
    orientation = "y",
    position = position_dodge(width = 0.45)
  ) +
  geom_point(size = 2.5, position = position_dodge(width = 0.45)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_color_manual(values = pal[c(2, 6)]) +
  labs(
    title = "Adjusted association with ECE seats per child",
    subtitle = "Estimated change for each 10 percentage-point increase; White is the reference composition",
    x = "Percent change",
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

pumas <- st_read("data/puma_nyc.geojson", quiet = TRUE)
pumas$puma <- sprintf("%04d", as.integer(pumas$puma))
map_dt <- merge(pumas, dt, by = "puma", all.x = TRUE)

map_access <- ggplot(map_dt) +
  geom_sf(aes(fill = occupied_seats_per_child), color = "white", linewidth = 0.12) +
  scale_fill_gradientn(
    colors = councildown::pal_nycc("cool"),
    labels = number_format(accuracy = 0.01),
    name = "Occupied seats\nper child"
  ) +
  labs(title = "Occupied ECE seats per preschool-age child") +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

map_plurality <- ggplot(map_dt) +
  geom_sf(aes(fill = plurality_group), color = "white", linewidth = 0.12) +
  scale_fill_manual(
    values = c(
      "White" = warm_pal[1],
      "Black" = warm_pal[2],
      "Hispanic" = warm_pal[3],
      "Asian" = warm_pal[4],
      "Other / 2+" = warm_pal[5]
    ),
    name = NULL,
    drop = TRUE
  ) +
  labs(title = "Largest racial/ethnic population group by PUMA") +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  )
```

::::{.cr-section .bg-city layout="overlay-center"}

::: {.title-box}

<p class="series-name">NYC COUNCIL RACIAL EQUITY SERIES</p>

# How Do Early Childhood Education Seats Vary by Neighborhood and Race? {.cr-title}

<p class="cr-subtitle">
An analysis of ECE seat availability, occupied seats, neighborhood racial and ethnic composition, and household income
</p>

<p class="image-credit">
Photo: <a href="https://www.flickr.com/photos/nyccouncil/53774000950/">New York City Council</a>
</p>

:::

::::

::: {.report-background}

::: {.key-findings}

<p class="key-findings-label">KEY FINDINGS</p>

::: {.key-finding}
<span class="finding-arrow">-></span>
<p>ECE access varies sharply across PUMAs: occupied seats range from `r number(min(dt$occupied_seats_per_child, na.rm = TRUE), accuracy = 0.01)` to `r number(max(dt$occupied_seats_per_child, na.rm = TRUE), accuracy = 0.01)` per preschool-age child.</p>
:::

::: {.key-finding}
<span class="finding-arrow">-></span>
<p>A 10 percentage-point increase in Asian population share is associated with `r number(largest_adjusted$percent_change, accuracy = 0.1)`% more occupied seats per preschool-age child after accounting for median household income.</p>
:::

::: {.key-finding}
<span class="finding-arrow">-></span>
<p>The adjusted models do not show a clear positive association for Hispanic population share; Black estimates are positive but less precise.</p>
:::

:::

::: {.report-layout}

::: {.report-navigation}

<p class="toc-label">In this report</p>

<nav>
<a href="#overview">Overview</a>
<a href="#what-we-found">What We Found</a>
<a href="#observed-disparities">Observed Disparities</a>
<a href="#adjusted-models">Adjusted Models</a>
<a href="#data-snapshot">Data Snapshot</a>
<a href="#policy-implications">Policy Implications</a>
<a href="#limitations">Limitations</a>
</nav>

:::

::: {.article-column}

## Overview {#overview}

Access to early childhood education (ECE) can depend on where a family lives. A shortage of seats in a neighborhood can make it harder for families to find care nearby, even when seats are available elsewhere in the city.

This one-pager examines whether ECE access varies by neighborhood racial and ethnic composition and household income. The outcomes are measured at the Public Use Microdata Area (PUMA) level as **licensed ECE seats per preschool-age child ages 3-4** and **occupied ECE seats per preschool-age child**.

The racial and ethnic categories match the housing quality analysis: **White**, **Black**, **Hispanic**, **Asian**, and **Other / 2+**. The ACS data build keeps `AIAN_NHPI` and `Other_2plus` as separate analytic groups and combines them as **Other / 2+** for public-facing descriptive tables and maps.

## What We Found {#what-we-found}

Neighborhoods with larger Asian population shares have the strongest positive adjusted association with ECE access. In the occupied-seat model, a 10 percentage-point increase in Asian population share is associated with **`r number(largest_adjusted$percent_change, accuracy = 0.1)`% more occupied seats per preschool-age child** after accounting for median household income.

Black population share is also positively associated with seats per child, but the adjusted estimates are smaller and less precise. Hispanic population share is not meaningfully associated with higher seat access after adjustment in these models.

## What disparities do we observe? {#observed-disparities}

Descriptively, ECE access varies sharply across PUMAs. The citywide mean is `r number(mean(dt$occupied_seats_per_child, na.rm = TRUE), accuracy = 0.01)` occupied seats per preschool-age child, while the PUMA-level range runs from `r number(min(dt$occupied_seats_per_child, na.rm = TRUE), accuracy = 0.01)` to `r number(max(dt$occupied_seats_per_child, na.rm = TRUE), accuracy = 0.01)`.

```{r}
#| label: map-access
#| fig-width: 7
#| fig-height: 5
map_access
```

Areas with different racial and ethnic plurality groups show different average levels of access, but these descriptive comparisons combine demographics, income, geography, enrollment preferences, and the historical location of ECE providers.

```{r}
#| label: map-plurality
#| fig-width: 7
#| fig-height: 5
map_plurality
```

```{r}
#| label: descriptive-table
desc_table <- composition_summary[
  ,
  .(
    `Largest group` = plurality_group,
    PUMAs = pumas,
    `Mean total seats per child` = ifelse(
      is.na(mean_total_seats_per_child),
      "n/a",
      number(mean_total_seats_per_child, accuracy = 0.01)
    ),
    `Mean occupied seats per child` = ifelse(
      is.na(mean_occupied_seats_per_child),
      "n/a",
      number(mean_occupied_seats_per_child, accuracy = 0.01)
    )
  )
]
knitr::kable(desc_table)
```

## Do those disparities persist after accounting for relevant factors? {#adjusted-models}

The adjusted models estimate log seats per child as a function of race/ethnicity composition and median household income. Coefficients are converted to percent changes, so each race/ethnicity estimate represents the modeled change associated with a 10 percentage-point increase in the population share. White is the omitted reference composition.

```{r}
#| label: adjusted-coef-plot
#| fig-width: 7
#| fig-height: 3.4
coef_plot
```

```{r}
#| label: adjusted-results-table
results_table <- race_results[
  ,
  .(
    Outcome = outcome,
    `Race/ethnicity share` = term_label,
    `Percent change` = number(percent_change, accuracy = 0.1),
    `95% CI` = paste0(
      number(percent_change_low, accuracy = 0.1),
      "% to ",
      number(percent_change_high, accuracy = 0.1),
      "%"
    ),
    `P-value` = fmt_p(p.value)
  )
]
knitr::kable(results_table)
```

The adjusted results suggest that the descriptive differences are not explained entirely by median household income. However, the model is ecological: it estimates associations across neighborhoods, not individual child participation or household-level access.

## Data Snapshot {#data-snapshot}

```{r}
#| label: data-snapshot-table
knitr::kable(snapshot)
```

**Data sources:** DOE ECE seat and vacancy data aggregated to PUMA, preschool-age child counts, and 2024 PUMA demographic and income measures in `data/lm_model_data.csv`; PUMA geography in `data/puma_nyc.geojson`.

## Policy Implications {#policy-implications}

Policy responses should focus on both **capacity** and **use**. Neighborhoods with low occupied seats per preschool-age child may need additional seats, outreach to families, language access, transportation support, or schedule models that better match caregiver needs. Areas where seats exist but occupancy is lower may require different interventions than areas with low total capacity.

Equity monitoring should track access by geography and racial/ethnic composition together, because citywide averages can obscure neighborhood-level mismatches between child population, available seats, and actual enrollment.

## Limitations {#limitations}

The analysis is PUMA-level and should not be interpreted as individual-level evidence. The regression models adjust for median household income but do not account for all relevant factors, including parental work schedules, program quality, language access, disability access, commute patterns, provider type, eligibility rules, or family preferences.

Seat counts and occupied seats are point-in-time administrative measures, and PUMAs can hide important variation within neighborhoods. The findings identify where disparities exist and whether they persist after a limited adjustment set; they do not establish why those disparities occur.

:::

:::

:::
