---
title: "Equity: ECE Seats"
author: "Brook Frye"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    df_print: paged
---

<!--
Steps:
1. Load total-seat model and map objects.
2. Plot seats against preschool-age children.
3. Show model results and diagnostics.
4. Render the total-seats page.
-->


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 6.5,
  fig.height = 3.8,
  out.width = "82%",
  fig.align = "center",
  dpi = 150
)
project_root <- getwd()
knitr::opts_knit$set(root.dir = project_root)
local({
  old_wd <- getwd()
  on.exit(setwd(old_wd), add = TRUE)
  setwd(project_root)
  source("00_load_dependencies.R")
  source("03_lm_mod_seats_demo_pums-2024.R")
  source("04_map_seats_child.R")
})
```

```{css, echo=FALSE}
body {
  font-size: 15px;
  line-height: 1.45;
}

.main-container {
  max-width: 920px;
}

h1, h2, h3 {
  margin-top: 1.35rem;
}

.table {
  width: auto !important;
  max-width: 760px;
  margin-left: auto;
  margin-right: auto;
  font-size: 0.88rem;
}

.table th,
.table td {
  padding: 5px 8px !important;
  white-space: normal !important;
}

.model-equation {
  max-width: 760px;
  margin: 0.75rem auto 1rem auto;
  padding: 0.75rem 1rem;
  border: 1px solid #d9e2ec;
  border-radius: 6px;
  background: #f8fafc;
  overflow-x: auto;
  font-size: 0.9rem;
}

.model-equation .MathJax_Display,
.model-equation mjx-container {
  margin: 0 !important;
  text-align: left !important;
}

.map-box {
  max-width: 760px;
  margin: 0 auto;
}

.map-box .leaflet,
.map-box .html-widget {
  height: 420px !important;
}

pre {
  max-width: 760px;
  margin-left: auto;
  margin-right: auto;
  font-size: 0.82rem;
  white-space: pre-wrap;
}
```

### Relationship between 3-4 year olds and total seats by PUMA

```{r seats-child-plot, fig.width=6.2, fig.height=3.6, out.width="78%"}
seats_child_p +
  labs(
    x = "Number of preschool-age children",
    y = "Total seats"
  ) +
  theme_minimal(base_size = 11)
```

### Model

We take the log of total seats per preschool-age child and regress it on race/ethnicity shares and median household income at the PUMA level.

This model asks: after accounting for the size of the preschool-age population, are neighborhood racial/ethnic makeup or household income associated with having more or fewer seats?

<div class="model-equation">
\[
\begin{aligned}
\log\left(\frac{\text{seats}}{\text{preschool-age children}}\right)
&= \beta_0 \\
&\quad + \beta_1(\text{Hispanic/Latino share}) \\
&\quad + \beta_2(\text{Black non-Hispanic share}) \\
&\quad + \beta_3(\text{Asian/API non-Hispanic share}) \\
&\quad + \beta_4(\text{median household income}) \\
&\quad + \epsilon
\end{aligned}
\]
</div>


- **Asian/API non-Hispanic population share has the strongest relationship with number of ECE seats.** A 10 % point increase in Asian/API NH population share is associated with ~20% more seats.
- **Black non-Hispanic population share is also positively associated with seats per child**
- **Hispanic/Latino population share is not meaningfully associated with seats per child in this model.**
- **Median household income is not strongly associated with seats per child after accounting for neighborhood racial/ethnic makeup**

```{r model-table}
knitr::kable(
  results_tab,
  caption = "Model results: percent change in seats per preschool-age child",
  align = c("l", "r", "r", "r", "r"),
  digits = 2,
  table.attr = "class='table table-striped table-hover table-condensed'"
)
```

```{r coefficient-plot, fig.width=6.4, fig.height=3.7, out.width="80%"}
coef_p +
  labs(title = "Neighborhood characteristics and seats per child") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(size = 12),
    plot.subtitle = element_text(size = 10),
    axis.text.y = element_text(size = 9)
  )
```

```{r map, out.width="100%"}
htmltools::div(
  class = "map-box",
  map
)
```

### Model Diagnostics

```{r}
pred_obs_p
resid_fitted_p
```

### Robustness Checks 

Dropping median HH income from model, we see that income attenuates the NH Black category, suggesting that part of the observed association between Black population share and seats/child is explained by differences in neighborhood income.For both models, Asian/API NH remains significant and suggests that in these neighborhoods with higher shares of Asian/API NH have more seats, regardless of median income. 

```{r}
kable(
  compare_table,
  caption = "Percent change in seats per preschool-age child, with and without income adj",
  align = c("l", "r", "r", "r", "r")
) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  add_header_above(
    c(
      " " = 1,
      "Race/ethnicity only model" = 2,
      "Race/ethnicity + income model" = 2
    )
  )

```
