---
output:
  pdf_document:
    toc: false
  html_document:
    theme: flatly
    toc: false
    df_print: paged
geometry: margin=0.7in
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 5.4,
  fig.height = 2.5,
  out.width = "82%",
  fig.align = "center",
  dpi = 150
)

source("2026/ECE/00_load_dependencies.R")
project_root <- getwd()
knitr::opts_knit$set(root.dir = project_root)



cool_pal <- councildown::pal_nycc("cool")


setDTthreads(1L)

fmt_p <- function(p) {
  fcase(
    is.na(p), NA_character_,
    p < 0.001, "<0.001",
    default = sprintf("%.3f", p)
  )
}

dtb <- fread(file.path(project_root, "2026/ECE/data/lm_model_data.csv"))
dtb[, puma := sprintf("%04d", as.integer(puma))]

dtb[
  ,
  `:=`(
    occupied_seats = total_seats - total_vacancy,
    occupied_seats_per_child = (total_seats - total_vacancy) / pre_sch_age
  )
]

total_model <- lm(
  log(total_seats / pre_sch_age) ~
    pct_hispanic10 +
    pct_black10 +
    pct_asian10 +
    median_inc_10k,
  data = dtb
)

occupied_model <- lm(
  log(occupied_seats_per_child) ~
    pct_hispanic10 +
    pct_black10 +
    pct_asian10 +
    median_inc_10k,
  data = dtb
)

prep_results <- function(model) {
  out <- as.data.table(tidy(model, conf.int = TRUE))
  
  out[
    ,
    `:=`(
      percent_change = 100 * (exp(estimate) - 1),
      percent_change_low = 100 * (exp(conf.low) - 1),
      percent_change_high = 100 * (exp(conf.high) - 1),
      term_label = fcase(
        term == "(Intercept)", "Intercept",
        term == "pct_hispanic10", "10 pp inc Hispanic/Latino",
        term == "pct_black10", "10 pp inc Black NH",
        term == "pct_asian10", "10 pp inc Asian/API NH",
        term == "median_inc_10k", "$10k inc Median hh income",
        default = term
      )
    )
  ]
  
  out
}

total_results <- prep_results(total_model)
occupied_results_keep <- prep_results(occupied_model)

clean_terms <- c(
  "pct_hispanic10",
  "pct_black10",
  "pct_asian10",
  "median_inc_10k"
)

coef_plot <- function(results, title, x_lab) {
  ggplot(
    results[term %in% clean_terms],
    aes(
      x = percent_change,
      y = reorder(term_label, percent_change)
    )
  ) +
    geom_vline(
      xintercept = 0,
      linetype = 2,
      colour = "grey60"
    ) +
    geom_errorbar(
      aes(
        xmin = percent_change_low,
        xmax = percent_change_high
      ),
      height = 0.15,
      orientation = "y"
    ) +
    geom_point(size = 2.3) +
    labs(
      title = title,
      subtitle = "Reference group: White non-Hispanic",
      x = x_lab,
      y = NULL
    ) +
    theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(size = 11),
      plot.subtitle = element_text(size = 9),
      axis.text.y = element_text(size = 8)
    )
}

total_coef_p <- coef_plot(
  total_results,
  "Total seats per child",
  "Percent change"
)

occupied_coef_p <- coef_plot(
  occupied_results_keep,
  "Occupied seats per child",
  "Percent change"
)

pumas <- st_read(file.path(project_root, "2026/ECE/data/puma_nyc.geojson"), quiet = TRUE)
pumas$puma <- sprintf("%04d", as.integer(pumas$puma))

occupied_map_dt <- merge(
  pumas,
  dtb[, .(puma, occupied_seats_per_child)],
  by = "puma",
  all.x = TRUE
)

occupied_map_p <- ggplot(occupied_map_dt) +
  geom_sf(
    aes(fill = occupied_seats_per_child),
    color = "white",
    linewidth = 0.12
  ) +
  scale_fill_gradientn(
    colors = cool_pal,
    labels = number_format(accuracy = 0.01),
    name = "Occupied seats\nper child"
  ) +
  labs(title = "Occupied ECE seats per preschool-age child") +
  theme_void(base_size = 10) +
  theme(
    plot.title = element_text(size = 11, hjust = 0.5),
    legend.position = "right"
  )

```

```{css, echo=FALSE}
body {
  font-size: 14px;
  line-height: 1.38;
}

.main-container {
  max-width: 860px;
}

h1 {
  font-size: 24px;
}

h2 {
  font-size: 18px;
  margin-top: 1.2rem;
}

.table {
  width: auto !important;
  max-width: 820px;
  margin-left: auto;
  margin-right: auto;
  font-size: 0.84rem;
}

.table th,
.table td {
  padding: 4px 7px !important;
  white-space: normal !important;
}
```

### **Early childhood education: How are demographics associated with total seats and occupied seats?**

**Brook Frye**

We explored neighborhood variation in DOE **early childhood education (ECE) seats per preschool-age child (ages 3–4)** and **occupied ECE seats per preschool-age child**. We estimated the relationship between the number of seats and number of occupied seats per child and neighborhood race/ethnicity shares, median household income.

The two models tell a consistent story. Asian/API non-Hispanic population share has the strongest association with early childhood education (ECE) access. **A 10 percentage-point increase in the Asian/API non-Hispanic population share is associated with approximately 20% more licensed seats per preschool-age child and 23% more occupied seats per preschool-age child.** This suggests a relationship between participation in ECE programs and number of seats per child in an area.

Black non-Hispanic population share is also positively associated with ECE capacity, but the association is attenuated after adjusting for median household income. Median household income has a slight negative association with seats, though the estimate is not significant. Hispanic/Latino population share is not associated with seats after adjustment.

## Coefficient Plots

```{r total-coef-plot, fig.width=6.2, fig.height=2.6, out.width="90%"}
total_coef_p
```

```{r occupied-coef-plot, fig.width=6.2, fig.height=2.6, out.width="90%"}
occupied_coef_p
```

## Occupied Seats Map

```{r occupied-map, fig.width=7.1, fig.height=5.2, out.width="95%"}
occupied_map_p
```
