---
title: "Equity: ECE Occupied Seats Per Child"
author: "Brook Frye"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    df_print: paged
---

<!--
Steps:
1. Load occupied-seat model and map objects.
2. Plot occupied seats against preschool-age children.
3. Show model results and diagnostics.
4. Render the occupied-seats page.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 6.5,
  fig.height = 3.8,
  out.width = "82%",
  fig.align = "center",
  dpi = 150
)
project_root <- getwd()
knitr::opts_knit$set(root.dir = project_root)
local({
  old_wd <- getwd()
  on.exit(setwd(old_wd), add = TRUE)
  setwd(project_root)
  source("00_load_dependencies.R")
  source("06_lm_mod_occupied_seats_pums-2024.R")
  source("07_map_occ_seats.R")
})

dtb <- fread(file.path(project_root, "data/lm_model_data.csv"))

dtb[
  ,
  `:=`(occupied_seats = total_seats - total_vacancy,
       occupied_seats_per_child = (total_seats - total_vacancy) /pre_sch_age)]

```

```{css, echo=FALSE}
body {
  font-size: 15px;
  line-height: 1.45;
}

.main-container {
  max-width: 920px;
}

h1, h2, h3 {
  margin-top: 1.35rem;
}

.table {
  width: auto !important;
  max-width: 760px;
  margin-left: auto;
  margin-right: auto;
  font-size: 0.88rem;
}

.table th,
.table td {
  padding: 5px 8px !important;
  white-space: normal !important;
}

.model-equation {
  max-width: 760px;
  margin: 0.75rem auto 1rem auto;
  padding: 0.75rem 1rem;
  border: 1px solid #d9e2ec;
  border-radius: 6px;
  background: #f8fafc;
  overflow-x: auto;
  font-size: 0.9rem;
}

.model-equation .MathJax_Display,
.model-equation mjx-container {
  margin: 0 !important;
  text-align: left !important;
}

.map-box {
  max-width: 760px;
  margin: 0 auto;
}

.map-box .leaflet,
.map-box .html-widget {
  height: 420px !important;
}

pre {
  max-width: 760px;
  margin-left: auto;
  margin-right: auto;
  font-size: 0.82rem;
  white-space: pre-wrap;
}
```

#### Relationship between 3-4 year olds and occupied seats by PUMA

```{r occupied-child-plot, fig.width=6.2, fig.height=3.6, out.width="78%"}
ggplot(
  dtb,
  aes(
    x = pre_sch_age,
    y = occupied_seats
  )
) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Preschool-age children",
    y = "Occupied seats"
  ) +
  theme_minimal(base_size = 11)
```

#### Occupied seats per child

Occupied seats per child is total seats minus vacant seats, divided by total preschool-age children.

#### Model

We take the log of occupied seats per preschool-age child and regress it on race/ethnicity shares and median household income at the PUMA level.

This model asks: after accounting for the size of the preschool-age population, are neighborhood racial/ethnic makeup or household income associated with having more or fewer occupied seats?

::: model-equation
$$
\begin{aligned}
\log\left(\frac{\text{licensed seats} - \text{vacant seats}}{\text{preschool age children}}\right)
&= \beta_0 \\
&\quad + \beta_1(\text{Hispanic/Latino share}) \\
&\quad + \beta_2(\text{Black non-Hispanic share}) \\
&\quad + \beta_3(\text{Asian/API non-Hispanic share}) \\
&\quad + \beta_4(\text{median household income}) \\
&\quad + \epsilon
\end{aligned}
$$
:::

-   **Asian/API non-Hispanic population share has the strongest relationship with occupied ECE seats.** A 10 percentage-point increase in Asian/API NH population share is associated with approximately 23% more occupied seats per preschool-age child.
-   **Black non-Hispanic population share is positively associated with occupied seats per child, but the association is not statistically significant after adjusting for neighborhood income.**
-   **Hispanic/Latino population share is not meaningfully associated with occupied seats per preschool-age child in this model.**
-   **Median household income is not strongly associated with occupied seats per preschool-age child** after accounting for neighborhood racial/ethnic composition.

```{r occupied-table}
knitr::kable(
  occupied_tab,
  caption = "Model results: percent change in occupied seats per preschool-age child",
  align = c("l", "r", "r", "r", "r"),
  digits = 2,
  table.attr = "class='table table-striped table-hover table-condensed'"
)
```

```{r occupied-coef-plot, fig.width=6.4, fig.height=3.7, out.width="80%"}
p_coef
```

#### Map of occupied seats by PUMA

```{r map, out.width="100%"}
occupied_map 
```

#### Model Diagnostics

```{r}
pred_obs_p
resid_fitted_p
```

#### Robustness Checks

-   **Black NH: Income explains very little of the relationship, but accounting for income increases uncertainty enough that the estimate is no longer significant.**
-   **The Hispanic/Latino association is attenuated after adjustment for income**, implying that income accounts for most of the modest association observed in the model.
-   **The Asian/API NH association remains large and highly statistically significant even after adjusting for income.**

```{r}
kable(
  compare_table,
  caption = "Percent change in occupied seats per preschool-age child, with and without income adj",
  align = c("l", "r", "r", "r", "r")
) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  add_header_above(
    c(
      " " = 1,
      "Race/ethnicity only model" = 2,
      "Race/ethnicity + income model" = 2
    )
  )

```
