# =============================================================================
# SCRIPT VERSION: pt8-ipw  (NB engine = withReplicates; + missingness diagnostics & IPW)
# If the knitted output does not print "[pt8-ipw]" near the top, you are
# running an older copy of this file.
# =============================================================================
# Controlled regression: UNSAFE_COND by race/ethnicity
# Howell-replication NYCHVS dataset
#
# Model:
#   UNSAFE_COND ~ RACE
#               + SUBSIDY
#               + OLDER_ADULT + SINGLE + CHILDREN
#               + YEARBUILT_ORD + UNITS_BIN + ROOMS_RAW
#
# Borough dummies excluded: VIF check confirmed rank deficiency of 4 due to
# collinearity with subsidy composition. YEARBUILT_ORD and UNITS_BIN serve
# as proxies for borough-level structural variation.
#
# CHANGE FROM PRIOR VERSION (pt3):
#   The prior script fit an unweighted-in-spirit glmmTMB NB model and then
#   hand-computed NYCHVS SDR replicate-weight SEs by refitting 80 times on
#   FW1..FW80. That approach produced correct point estimates and correct
#   SEs, but any likelihood-ratio-based test built on top of it (car::Anova
#   Type-II tests, and the NB-vs-ZINB LRT) was invalid: passing a survey
#   weight into glmmTMB's likelihood as if it were a frequency weight
#   inflates the effective sample size the likelihood "sees," which inflates
#   every LRT chi-square and collapses p-values regardless of the true
#   effect size. A parallel check with survey::svrepdesign() + svyglm()
#   confirmed the manual SDR math was correct (SE ratios 0.91-1.05 vs. a
#   quasipoisson cross-check) but also confirmed the LR-based joint tests
#   were unusable (car::Anova chi-squares in the thousands vs. regTermTest
#   F-stats in the single-to-low-double digits on the same data).
#
#   This version replaces that whole two-model workaround with a single,
#   properly-specified survey-weighted model, fit directly against the SDR
#   replicate-weight design:
#     - If the data are overdispersed (same ratio check as before):
#         sjstats::svyglm.nb() -- a true survey-weighted negative binomial,
#         fit via survey::svymle() with glm.nb() starting values (Lumley
#         2010, Complex Surveys, Appendix E).
#     - If svyglm.nb() fails to converge, or if the data are NOT
#       overdispersed: svyglm() with quasipoisson (overdispersed but no
#       stable NB fit) or poisson (not overdispersed) as appropriate.
#   Coefficients, SEs, AND joint significance tests now all come from one
#   correctly-specified survey model instead of mixing an NB point estimate
#   with a quasipoisson test statistic.
#
# Variance estimation:
#   NYCHVS SDR replicate weights (FW1..FW80) are handled natively by
#   survey::svrepdesign(), using the same (4/80) scaling constant as the
#   manual loop in the prior version, so results are on the same basis.
#
# Zero-inflation:
#   Dropped from this version. DHARMa's testZeroInflation() on the plain NB
#   model previously came back non-significant in all three samples
#   (p = 0.074 / 1.000 / 0.304), and the ZINB coefficients were nearly
#   identical to the NB's -- so the earlier "ZINB fits significantly
#   better" LRT result was itself an artifact of the same weight-inflation
#   bug, not evidence of real excess zeros. No ZINB step is needed.
#
# Outputs:
#   controlled_svy_unsafe_cond_results.csv   full coefficient table
#   dharma_diagnostics_svy.pdf                residual diagnostics (fit on an
#                                              equivalent unweighted-design
#                                              MASS::glm.nb / glm(poisson)
#                                              model -- see Section 10 note)
# =============================================================================


# -----------------------------------------------------------------------------
# 0. Packages and environment setup
# -----------------------------------------------------------------------------
```{r}
# Create new environment at "equity_projects/2026/housing_quality/"
setwd("./2026/housing_quality")
source("renv/activate.R")

library(tidyverse)
library(survey)
library(MASS)
library(DHARMa)
library(car)
library(broom)

# NOTE: glmmTMB is intentionally no longer a required package for the
# reported model. If you still want a glmmTMB fit for cross-checking,
# install/load it separately -- it is not used below.
```



# -----------------------------------------------------------------------------
# 1. Configuration
# -----------------------------------------------------------------------------
```{r}
INPUT_CSV  <- "./data/processed/howell_replication_nychvs.csv"
OUTPUT_CSV <- "./data/processed/controlled_svy_unsafe_cond_results.csv"
DIAG_PDF   <- "./visuals/dharma_diagnostics_svy.pdf"

N_REPLICATES           <- 80L
SDR_SCALE              <- 4.0 / N_REPLICATES
OVERDISPERSION_THRESH  <- 1.5
STANDARDIZE            <- FALSE   # set TRUE to z-score continuous predictors

OUTCOME_COL   <- "UNSAFE_COND"
OUTCOME_LABEL <- "Unsafe conditions (count, 0-6)"

RACE_LEVELS <- c("White", "Black", "Hispanic", "Asian",
                 "AIAN_NHPI", "Other_2plus")

SUBSIDY_LEVELS_ALL   <- c("Unsubsidized", "Voucher", "Other_subsidy")
SUBSIDY_LEVELS_SUB   <- c("Unsubsidized", "Voucher", "Other_subsidy")
SUBSIDY_LEVELS_UNSUB <- "Unsubsidized"

REGSTATUS_LEVELS <- c("Market", "Rent_stabilized", "Rent_controlled",
                      "Other_regulated", "Public_NYCHA")

CONTINUOUS_PREDS <- c("YEARBUILT_ORD", "UNITS_BIN", "ROOMS_RAW")

ALL_PREDICTORS <- c(
  "RACE", "SUBSIDY", "REGSTATUS", "OLDER_ADULT", "SINGLE", "CHILDREN",
  "YEARBUILT_ORD", "UNITS_BIN", "ROOMS_RAW"
)
```



# -----------------------------------------------------------------------------
# 2. Load and prepare
# -----------------------------------------------------------------------------
```{r}
df_raw <- read_csv(INPUT_CSV, show_col_types = FALSE)
cat(sprintf("Loaded %s rows.\n\n", format(nrow(df_raw), big.mark = ",")))

rep_cols <- paste0("FW", seq_len(N_REPLICATES))
has_rep  <- all(rep_cols %in% names(df_raw))

if (has_rep) {
  cat("Detected FW1..FW80 -- using NYCHVS SDR replicate-weight design.\n\n")
} else {
  stop("FW1..FW80 not detected -- this script requires replicate weights ",
       "to build the survey design. Falling back to model-based SEs is not ",
       "implemented in this version; use the pt3 script's fallback path if ",
       "replicate weights are unavailable.")
}

df <- df_raw |>
  mutate(
    IS_PUBLIC     = REGSTATUS == "Public_NYCHA",
    IS_SUBSIDIZED = SUBSIDY %in% c("Voucher", "Other_subsidy") | IS_PUBLIC,
    RACE      = factor(RACE,      levels = RACE_LEVELS),
    SUBSIDY   = factor(SUBSIDY,   levels = SUBSIDY_LEVELS_ALL),
    REGSTATUS = factor(REGSTATUS, levels = REGSTATUS_LEVELS)
  )
```



# -----------------------------------------------------------------------------
# 3. Overdispersion check (same logic as before -- decides NB vs. Poisson path)
# -----------------------------------------------------------------------------
```{r}
uc_mean <- mean(df[[OUTCOME_COL]], na.rm = TRUE)
uc_var  <- var(df[[OUTCOME_COL]],  na.rm = TRUE)
ratio   <- uc_var / uc_mean

cat(strrep("=", 60), "\n")
cat(sprintf("OVERDISPERSION CHECK  (%s, full dataset)\n", OUTCOME_COL))
cat(strrep("=", 60), "\n")
cat(sprintf("  mean : %.4f\n  var  : %.4f\n  ratio: %.4f\n", uc_mean, uc_var, ratio))

use_nb <- ratio > OVERDISPERSION_THRESH
cat(sprintf("  -> Overdispersed: %s (threshold=%.1f)\n",
            use_nb, OVERDISPERSION_THRESH))
cat(sprintf("  -> Will attempt: %s\n\n",
            if (use_nb) "svymle NB [fallback: svyglm quasipoisson]"
            else        "svyglm() poisson"))
```



# -----------------------------------------------------------------------------
# 4. Collinearity check (VIF on a Poisson GLM -- unchanged diagnostic)
# -----------------------------------------------------------------------------
```{r}
cat(strrep("=", 60), "\n")
cat("COLLINEARITY CHECK (VIF, all-renters sample)\n")
cat(strrep("=", 60), "\n")

df_vif <- df |>
  filter(!is.na(UNSAFE_COND), !is.na(FW)) |>
  droplevels()

df_vif$SUBSIDY <- relevel(
  factor(df_vif$SUBSIDY,
         levels = intersect(SUBSIDY_LEVELS_ALL, unique(as.character(df_vif$SUBSIDY)))),
  ref = "Unsubsidized"
)
df_vif$REGSTATUS <- relevel(
  factor(df_vif$REGSTATUS,
         levels = intersect(REGSTATUS_LEVELS, unique(as.character(df_vif$REGSTATUS)))),
  ref = "Market"
)

vif_fit <- glm(
  UNSAFE_COND ~ RACE + SUBSIDY + REGSTATUS + OLDER_ADULT + SINGLE + CHILDREN +
    YEARBUILT_ORD + UNITS_BIN + ROOMS_RAW,
  data    = df_vif,
  family  = poisson(),
  weights = FW
)
vif_vals <- tryCatch(car::vif(vif_fit), error = function(e) {
  cat("  (VIF computation failed:", conditionMessage(e), ")\n")
  NULL
})
if (!is.null(vif_vals)) {
  if (is.matrix(vif_vals)) {
    vif_df <- as.data.frame(vif_vals) |>
      rownames_to_column("term") |>
      rename(GVIF = 1, Df = 2, adj_GVIF = 3)
    cat("  (Generalized VIF -- adj_GVIF = GVIF^(1/(2*Df)), comparable to sqrt(VIF))\n")
    print(vif_df |> arrange(desc(adj_GVIF)), digits = 3)
  } else {
    vif_df <- tibble(term = names(vif_vals), VIF = unname(vif_vals))
    print(vif_df |> arrange(desc(VIF)), digits = 3)
  }
}
cat("\n")
```



# -----------------------------------------------------------------------------
# 5. Helper: prepare analytic subset (unchanged from prior version)
# -----------------------------------------------------------------------------
```{r}
prep_subset <- function(data, subsidy_levels, standardize = FALSE) {
  needed <- c(ALL_PREDICTORS, OUTCOME_COL, "FW")

  sub <- data |>
    drop_na(any_of(needed)) |>
    mutate(RACE = factor(RACE, levels = RACE_LEVELS))

  present_sub <- intersect(subsidy_levels,
                           unique(as.character(sub$SUBSIDY)))
  present_reg <- intersect(REGSTATUS_LEVELS,
                           unique(as.character(sub$REGSTATUS)))

  if ("Public_NYCHA" %in% present_reg) {
    is_pub <- sub$REGSTATUS == "Public_NYCHA"
    sub_of_pub <- unique(as.character(sub$SUBSIDY[is_pub]))
    redundant <- length(sub_of_pub) == 1 &&
      all(sub$REGSTATUS[sub$SUBSIDY == sub_of_pub] == "Public_NYCHA")
    if (redundant) {
      present_reg <- setdiff(present_reg, "Public_NYCHA")
      if (length(present_reg) == 0) {
        reg_ref <- "Market"
        present_reg <- "Market"
      } else {
        reg_ref <- present_reg[1]
      }
      sub <- sub |>
        mutate(REGSTATUS = if_else(REGSTATUS == "Public_NYCHA",
                                   reg_ref, as.character(REGSTATUS)))
      message(sprintf(
        paste0("  [prep_subset] REGSTATUS=='Public_NYCHA' is collinear with ",
               "SUBSIDY=='%s' in this subset; dropping the REGSTATUS public ",
               "level (public rows folded into REGSTATUS reference '%s'). ",
               "Public housing is carried by the SUBSIDY reference here."),
        sub_of_pub, reg_ref))
    }
  }

  sub <- sub |>
    mutate(
      SUBSIDY   = factor(as.character(SUBSIDY),   levels = present_sub),
      REGSTATUS = factor(as.character(REGSTATUS), levels = present_reg)
    )

  if (standardize) {
    sub <- sub |>
      mutate(across(all_of(CONTINUOUS_PREDS),
                    ~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE)))
  }
  sub
}
```



# -----------------------------------------------------------------------------
# 6. Build the model formula for a given subset (unchanged level-count gating)
# -----------------------------------------------------------------------------
```{r}
build_formula <- function(data) {
  n_sub_levels <- nlevels(droplevels(data$SUBSIDY))
  n_reg_levels <- nlevels(droplevels(data$REGSTATUS))
  terms <- c("RACE")
  if (n_sub_levels > 1) terms <- c(terms, "SUBSIDY")
  if (n_reg_levels > 1) terms <- c(terms, "REGSTATUS")
  terms <- c(terms, "OLDER_ADULT", "SINGLE", "CHILDREN",
             "YEARBUILT_ORD", "UNITS_BIN", "ROOMS_RAW")
  list(
    fml   = as.formula(paste("UNSAFE_COND ~", paste(terms, collapse = " + "))),
    terms = terms
  )
}
```



# -----------------------------------------------------------------------------
# 6b. Missing-outcome diagnostics and IPW sensitivity analysis
# -----------------------------------------------------------------------------
# UNSAFE_COND is NA wherever HPROBCOUNT == -1 ("not computed" because some
# component item was missing): 339 of 3,540 rows in the full file (9.6%
# unweighted, 10.7% weighted). The main models are complete-case, so we need to
# (a) show whether that missingness is ignorable and (b) quantify how much the
# estimates would move if we corrected for it.
#
# The relevant distinction:
#   MCAR  -- missingness unrelated to anything. Complete-case is unbiased.
#   MAR   -- missingness depends on OBSERVED covariates only. Complete-case is
#            still unbiased for a regression when the missingness is in the
#            OUTCOME and depends only on covariates already in the model.
#   MNAR  -- missingness depends on the unobserved outcome itself. Nothing here
#            can detect or fix that; it is a substantive judgement about the
#            instrument, and should be stated as a limitation.
#
# test_missingness() reports, per predictor block, both the marginal
# association with missingness and the association conditional on the rest of
# the model RHS. The conditional column is the one that matters: if RACE is
# marginally associated with missingness but NOT conditionally, the complete-
# case race coefficients are protected under MAR.
# -----------------------------------------------------------------------------
```{r}
test_missingness <- function(data, fml) {
  yv <- all.vars(fml)[1]
  d  <- data
  d$.RESP <- !is.na(d[[yv]])
  n_miss  <- sum(!d$.RESP)

  if (n_miss == 0) {
    cat("  [missingness] no missing outcome in this subset.\n")
    return(invisible(NULL))
  }

  cat(sprintf(
    "  [missingness] %d of %d rows (%.1f%% unweighted, %.1f%% weighted) lack %s\n",
    n_miss, nrow(d), 100 * mean(!d$.RESP),
    100 * sum(d$FW[!d$.RESP]) / sum(d$FW), yv))

  preds <- attr(terms(fml), "term.labels")

  # Marginal: is missingness related to this variable at all?
  marg <- do.call(rbind, lapply(preds, function(v) {
    x <- d[[v]]
    p <- tryCatch({
      if (is.factor(x) || is.character(x)) {
        suppressWarnings(chisq.test(table(x, d$.RESP))$p.value)
      } else {
        suppressWarnings(t.test(x[!d$.RESP], x[d$.RESP])$p.value)
      }
    }, error = function(e) NA_real_)
    tibble(term = v, marginal_p = p)
  }))

  # Conditional: does it still predict missingness given the whole RHS?
  pm <- suppressWarnings(
    glm(reformulate(preds, response = ".RESP"), data = d, family = binomial))
  cf <- coef(pm); V <- vcov(pm)
  cond <- do.call(rbind, lapply(preds, function(v) {
    idx <- grep(paste0("^", v), names(cf))
    idx <- idx[is.finite(cf[idx])]
    if (!length(idx)) return(NULL)
    b  <- cf[idx]; Vb <- V[idx, idx, drop = FALSE]
    st <- tryCatch(as.numeric(t(b) %*% solve(Vb) %*% b),
                   error = function(e) NA_real_)
    tibble(term = v,
           conditional_p = if (is.na(st)) NA_real_ else
             pchisq(st, df = length(idx), lower.tail = FALSE))
  }))

  out <- left_join(marg, cond, by = "term") |> arrange(conditional_p)
  print(out |> mutate(across(where(is.numeric), \(x) signif(x, 3))), n = Inf)
  cat("  (marginal_p: related to missingness at all?  conditional_p: still\n")
  cat("   predictive given the rest of the RHS?  Small marginal_p rules out\n")
  cat("   MCAR. Complete-case remains unbiased under MAR -- what matters is\n")
  cat("   that the KEY exposure is not conditionally predictive.)\n")
  invisible(out)
}

# -----------------------------------------------------------------------------
# IPW sensitivity refit: model the probability of having an observed outcome
# (the "response propensity") on the model covariates, then reweight complete
# cases by FW / p_response. This restores the population totals that the
# complete-case analysis loses, and gives an estimate that is consistent under
# MAR even if the complete-case one were not. Both the full-sample weight and
# all 80 replicate weights are scaled identically, so the replicate variance
# estimator (4/80 sum (theta_r - theta_0)^2) still applies unchanged.
# -----------------------------------------------------------------------------
ipw_refit <- function(data, fml) {
  yv    <- all.vars(fml)[1]
  preds <- attr(terms(fml), "term.labels")

  # Restrict to complete COVARIATES first: the propensity model can only be
  # fitted and predicted where the RHS is observed, and NA propensities would
  # poison the weights and break svrepdesign().
  covar_ok   <- stats::complete.cases(data[, all.vars(fml)[-1], drop = FALSE])
  n_cov_drop <- sum(!covar_ok)
  d          <- data[covar_ok, , drop = FALSE]
  d$.RESP    <- !is.na(d[[yv]])
  if (all(d$.RESP)) return(NULL)

  pm <- suppressWarnings(
    glm(reformulate(preds, response = ".RESP"), data = d, family = binomial))

  cc <- d[d$.RESP, , drop = FALSE]
  p  <- stats::predict(pm, newdata = cc, type = "response")
  p  <- pmax(p, 0.05)          # trim to avoid extreme inflation factors

  cc$FW <- cc$FW / p
  for (rc in rep_cols) cc[[rc]] <- cc[[rc]] / p
  cc <- droplevels(cc)

  des <- build_svy_design(cc)
  ww  <- as.numeric(stats::weights(des, type = "sampling"))
  cc$.ipwwt <- ww / mean(ww)
  m0  <- suppressWarnings(MASS::glm.nb(fml, data = cc, weights = .ipwwt))

  wr <- survey::withReplicates(des, function(w, data) {
    data$.nbwt <- w / mean(w)
    stats::coef(suppressWarnings(MASS::glm.nb(fml, data = data, weights = .nbwt)))
  })

  list(coefficients = stats::coef(wr),
       se           = survey::SE(wr),
       theta        = as.numeric(m0$theta),
       p_range      = range(p),
       n_cov_drop   = n_cov_drop,
       pop_cc       = sum(d$FW[d$.RESP]),
       pop_ipw      = sum(cc$FW))
}

# Side-by-side comparison of the main (complete-case) fit and the IPW refit.
compare_ipw <- function(main_fit, ipw, blocks = c("RACE", "SUBSIDY", "REGSTATUS")) {
  if (is.null(ipw)) {
    cat("  (no missing outcome in this subset -- IPW sensitivity not needed)\n")
    return(invisible(NULL))
  }
  cat(sprintf(
    "  IPW refit: theta=%.3f | response p in [%.3f, %.3f] | weighted N %s -> %s\n",
    ipw$theta, ipw$p_range[1], ipw$p_range[2],
    format(round(ipw$pop_cc), big.mark = ","),
    format(round(ipw$pop_ipw), big.mark = ",")))

  cm <- stats::coef(main_fit); sm <- main_fit$se
  ci <- ipw$coefficients;      si <- ipw$se
  keep <- intersect(names(cm), names(ci))
  keep <- keep[grepl(paste0("^(", paste(blocks, collapse = "|"), ")"), keep)]
  if (!length(keep)) return(invisible(NULL))

  tab <- tibble(
    term      = keep,
    cc_coef   = unname(cm[keep]),
    cc_se     = unname(sm[keep]),
    ipw_coef  = unname(ci[keep]),
    ipw_se    = unname(si[keep])
  ) |>
    mutate(diff      = ipw_coef - cc_coef,
           diff_in_se = diff / cc_se)

  print(tab |> mutate(across(where(is.numeric), \(x) round(x, 4))), n = Inf)
  cat("  (diff_in_se expresses the shift in complete-case SE units; |values|\n")
  cat("   well under 1 mean the IPW correction is substantively immaterial.)\n")
  invisible(tab)
}
```



# -----------------------------------------------------------------------------
# 7. Build the SDR replicate-weight survey design
# -----------------------------------------------------------------------------
# type = "other" tells survey these are pre-scaled generic replicates; scale
# and rscales below reproduce the (4/80) NYCHVS SDR constant natively, so the
# design-based variance matches what the manual SDR loop computed in the
# prior version (independently validated: SE ratio range 0.906-1.046).
# -----------------------------------------------------------------------------
```{r}
build_svy_design <- function(data) {
  svrepdesign(
    data              = data,
    weights           = ~FW,
    repweights        = data[, rep_cols],
    type              = "other",
    scale             = SDR_SCALE,
    rscales           = rep(1, N_REPLICATES),
    combined.weights  = TRUE,
    mse               = TRUE
  )
}
```



# -----------------------------------------------------------------------------
# 8. Fit the survey-weighted model, choosing family based on diagnostics
# -----------------------------------------------------------------------------
# Decision logic:
#   use_nb == TRUE  -> try sjstats::svyglm.nb() (true survey-weighted NB).
#                       If it errors or fails to converge, fall back to
#                       svyglm(..., family = quasipoisson(link="log")),
#                       which still corrects SEs for overdispersion even
#                       though it doesn't model the NB variance function.
#   use_nb == FALSE -> svyglm(..., family = poisson(link="log")); no
#                      overdispersion correction needed by the ratio check.
#
# NOTE: sjstats::svyglm.nb() is less battle-tested than glmmTMB and can be
# sensitive to starting values/separation in small strata (e.g. the
# Unsubsidized sample, where SUBSIDY drops out entirely). Convergence is
# checked explicitly; check the console messages on first run for each
# sample before trusting the reported family.
# -----------------------------------------------------------------------------
```{r}
# "scaled.weights" is created inside the survey design via update() and then
# referenced by name in glm.nb(weights = scaled.weights); declaring it here
# just silences the "no visible binding" NOTE.
utils::globalVariables("scaled.weights")

# ---------------------------------------------------------------------------
# Direct survey-weighted negative binomial via survey::svymle().
#
# This replaces sjstats::svyglm.nb(), which is unreliable to depend on across
# sjstats/survey versions (deprecated/removed export -> errors identically on
# every sample, which is exactly the "NB fails on all samples" symptom). The
# code below is a faithful vendored copy of sjstats' own implementation
# (strengejacke/sjstats, R/svyglmnb.R), which is itself just the Lumley (2010,
# Appendix E) recipe: svymle() with an NB2 log-likelihood, analytic score, and
# glm.nb() starting values. Vendoring it removes the package dependency while
# keeping the exact, tested likelihood/score -- including the analytic score,
# which is what lets svymle return the robust (sandwich) SEs we want from the
# replicate-weight design.
#
# Parameterization matches sjstats exactly: parameters are (theta, eta) in
# that order; theta is the NB size parameter (NOT log theta); mu = exp(eta).
# ---------------------------------------------------------------------------

# NB2 log-likelihood (per observation). The (y == 0) guard avoids log(0) when
# mu is used inside log(mu + (y==0)); it contributes 0 to the y*log(mu) term.
nb_loglik <- function(y, theta, eta) {
  mu <- exp(eta)
  lgamma(theta + y) - lgamma(theta) - lgamma(y + 1) +
    theta * log(theta) + y * log(mu + (y == 0)) -
    (theta + y) * log(theta + mu)
}
# d loglik / d eta
nb_deta <- function(y, theta, eta) {
  mu  <- exp(eta)
  dmu <- y / mu - (theta + y) / (theta + mu)
  dmu * mu
}
# d loglik / d theta
nb_dtheta <- function(y, theta, eta) {
  mu <- exp(eta)
  digamma(theta + y) - digamma(theta) + log(theta) + 1 -
    log(theta + mu) - (y + theta) / (mu + theta)
}
# score: columns must line up with the (theta, eta) parameter order
nb_score <- function(y, theta, eta) {
  cbind(nb_dtheta(y, theta, eta), nb_deta(y, theta, eta))
}

svy_nb_fit <- function(fml, design) {
  # ---------------------------------------------------------------------------
  # Survey-weighted negative binomial for a REPLICATE-WEIGHT design.
  #
  # Why not svymle(): survey::svymle() hard-requires inherits(design,
  # "survey.design") and has no method for svyrep.design objects. NYCHVS uses
  # replicate weights, so build_svy_design() returns an svyrep.design -- which
  # svymle() rejects outright ("design is not a survey.design"). This is also
  # why sjstats::svyglm.nb() could never have worked here: it wraps svymle().
  #
  # The correct tool is withReplicates(), survey's designated mechanism for
  # applying an arbitrary estimator across replicate weights. It computes
  # exactly the NYCHVS successive-differences estimator from the weights guide,
  #     v(theta) = (4/80) * sum_r (theta_r - theta_0)^2,
  # given scale = 4/80 and rscales = rep(1, 80) as set in build_svy_design().
  # Verified numerically against a hand-coded implementation of that formula
  # (agreement to ~5e-17).
  #
  # Point estimate: full-sample-weighted NB MLE (weights normalised to mean 1,
  # which leaves the coefficients unchanged but keeps glm.nb well-scaled).
  # SEs: replicate-based, per the guide.
  # ---------------------------------------------------------------------------
  fit_one <- function(w, data) {
    data$.nbwt <- w / mean(w)
    m <- suppressWarnings(
      MASS::glm.nb(fml, data = data, weights = .nbwt))
    stats::coef(m)
  }

  dat <- design$variables

  # Full-sample point estimate + theta (kept for reporting).
  # NOTE: weights(design) on an svyrep.design returns the FULL n x 80 REPLICATE
  # MATRIX, not the sampling weights. Passing that to glm.nb silently supplies
  # a weight vector 80x too long -> NAs -> stop("NAs in V(mu)"). Always request
  # type = "sampling" here to get the FW vector.
  dat$.fullwt <- {
    ww <- stats::weights(design, type = "sampling")
    as.numeric(ww) / mean(as.numeric(ww))
  }
  m0    <- suppressWarnings(
    MASS::glm.nb(fml, data = dat, weights = .fullwt))
  theta <- as.numeric(m0$theta)

  wr <- survey::withReplicates(design, fit_one)

  rval <- list(
    coefficients = stats::coef(wr),
    se           = survey::SE(wr),
    theta        = theta,
    glmnb        = m0,
    withrep      = wr
  )
  class(rval) <- "svynb_rep"
  rval
}

# Extractors so downstream code can treat this like any other fit.
coef.svynb_rep  <- function(object, ...) object$coefficients
vcov.svynb_rep  <- function(object, ...) stats::vcov(object$withrep)
SE.svynb_rep    <- function(object, ...) object$se


fit_survey_model <- function(data, use_nb) {
  fb     <- build_formula(data)
  fml    <- fb$fml
  terms  <- fb$terms

  # Drop incomplete cases BEFORE building the design. UNSAFE_COND has NAs in
  # the source extract (339 in the full file) and SINGLE has some too. If they
  # are left in, the design carries rows the model cannot use, and any
  # downstream row-subsetting desynchronises design$variables from the
  # replicate-weight matrix (this is what produced "(subscript) logical
  # subscript too long"). Filtering here keeps the design, the weights, and the
  # 80 replicate columns row-aligned by construction.
  n_before <- nrow(data)
  data     <- data[stats::complete.cases(data[, all.vars(fml), drop = FALSE]), ]
  data     <- droplevels(data)
  n_drop   <- n_before - nrow(data)
  if (n_drop > 0) {
    cat(sprintf("  [design] dropped %d incomplete row(s); n=%d used for modelling\n",
                n_drop, nrow(data)))
  }

  des    <- build_svy_design(data)

  family_used <- NA_character_
  fit         <- NULL
  fit_error   <- NULL

  if (use_nb) {
    fit_warnings <- character(0)
    fit <- tryCatch(
      withCallingHandlers(
        svy_nb_fit(fml, des),
        warning = function(w) {
          fit_warnings <<- c(fit_warnings, conditionMessage(w))
          invokeRestart("muffleWarning")
        }
      ),
      error = function(e) { fit_error <<- conditionMessage(e); NULL }
    )

    if (!is.null(fit)) {
      family_used <- sprintf("NB via withReplicates (theta=%.3f)", fit$theta)
      if (length(fit_warnings) > 0) {
        cat(sprintf(
          "  NB fit converged with %d warning(s) (non-fatal): %s\n",
          length(fit_warnings),
          paste(unique(fit_warnings), collapse = "; ")))
      }
    } else {
      # Surface to stdout (cat), not message(), so the reason is visible in
      # knitted output instead of being hidden on stderr.
      cat(sprintf(
        "  NB fit FAILED (%s)\n  -- falling back to svyglm quasipoisson.\n",
        fit_error))
    }
  }

  if (is.null(fit)) {
      if (use_nb) {
        fit         <- svyglm(fml, design = des, family = quasipoisson(link = "log"))
        family_used <- "svyglm quasipoisson (NB fallback)"
      } else {
        fit         <- svyglm(fml, design = des, family = poisson(link = "log"))
        family_used <- "svyglm poisson"
      }
    }

  list(fit = fit, design = des, terms = terms, family_used = family_used,
       data = data, fml = fml)
}
```



# -----------------------------------------------------------------------------
# 9. Generic coefficient/SE extraction
# -----------------------------------------------------------------------------
# Works across svyglm.nb (svymle-based) and svyglm (quasipoisson/poisson)
# objects without assuming a specific S3 method exists for each. Tries the
# most specific approach first and falls back progressively.
# -----------------------------------------------------------------------------
```{r}
get_model_coefs <- function(fit) {
  # NB fits from svy_nb_fit() carry coefficients and replicate-based SEs
  # directly; handle them first.
  if (inherits(fit, "svynb_rep")) {
    return(tibble(term = names(fit$coefficients),
                  coef = unname(fit$coefficients),
                  se   = unname(fit$se)))
  }

  out <- tryCatch({
    s <- summary(fit)
    cm <- if (!is.null(s$coefficients)) s$coefficients else coef(s)
    tibble(
      term = rownames(cm),
      coef = cm[, 1],
      se   = cm[, 2]
    )
  }, error = function(e) NULL)

  if (is.null(out)) {
    out <- tryCatch({
      cf <- coef(fit)
      se <- sqrt(diag(vcov(fit)))
      tibble(term = names(cf), coef = unname(cf), se = unname(se[names(cf)]))
    }, error = function(e) {
      stop("Could not extract coefficients/SEs from fit: ", conditionMessage(e))
    })
  }

  # svymle NB fits name parameters by block: the dispersion parameter as
  # "theta.(Intercept)" and the mean model as "eta.<term>". Drop the
  # dispersion row (anchored to the "theta." prefix so it can't collide with a
  # real predictor that happens to contain "theta") and strip the "eta."
  # prefix so downstream code (section labels, joint-test greps like "^RACE")
  # sees bare term names identical to what svyglm/glm produce. This is a no-op
  # for plain svyglm objects, whose names have neither prefix.
  out <- out |>
    filter(!grepl("^theta\\.", term)) |>
    mutate(term = sub("^eta\\.", "", term))
  out
}
```



# -----------------------------------------------------------------------------
# 10. Generic joint (block-level) significance test
# -----------------------------------------------------------------------------
# regTermTest() is documented for svyglm objects but is NOT guaranteed to
# support svyglm.nb/svymle objects. Rather than depend on that, this uses a
# model-agnostic multivariate Wald test built directly from coef()/vcov():
# for a predictor block (e.g. all "RACE*" dummy columns), it tests the joint
# null that all of that block's coefficients are 0, using the block's
# coefficient sub-vector and the corresponding sub-matrix of vcov(fit).
# This works identically regardless of which family/engine produced the fit.
#
# For plain svyglm objects (quasipoisson/poisson fallback), regTermTest() is
# ALSO attempted as a secondary cross-check, since it is documented to work
# there and gives an F-test with design degrees of freedom rather than a
# chi-square.
# -----------------------------------------------------------------------------
```{r}
joint_wald_test <- function(fit, term_prefix) {
  cf  <- coef(fit)
  V   <- vcov(fit)
  # svymle NB coefficients are named "eta.<term>"; svyglm/glm names are bare.
  # Match the block prefix optionally preceded by "eta." so this works for
  # both engines. The vcov row/col order matches names(cf), so indexing by
  # the same idx into V is valid in both cases.
  idx <- grep(paste0("^(eta\\.)?", term_prefix), names(cf))
  if (length(idx) == 0) return(NULL)

  b   <- cf[idx]
  Vb  <- V[idx, idx, drop = FALSE]
  stat <- as.numeric(t(b) %*% solve(Vb) %*% b)
  df   <- length(idx)
  p    <- pchisq(stat, df = df, lower.tail = FALSE)

  tibble(term = term_prefix, wald_chisq = stat, df = df, p = p)
}

run_joint_tests <- function(svy_obj) {
  # Use the same top-level predictor names used to build the formula, not a
  # regex guess -- this avoids splitting multi-level factors incorrectly.
  block_terms <- svy_obj$terms

  wald_rows <- lapply(block_terms, function(t) joint_wald_test(svy_obj$fit, t))
  wald_df <- bind_rows(wald_rows)

  # Secondary cross-check via regTermTest(), only meaningful for plain
  # svyglm (quasipoisson/poisson) objects. svymle NB fits are class "svymle",
  # not "svyglm", so the inherits() test already excludes them; regTermTest is
  # skipped for those and only the model-agnostic Wald table is reported.
  is_plain_svyglm <- inherits(svy_obj$fit, "svyglm") &&
    !grepl("svymle", svy_obj$family_used)

  if (is_plain_svyglm) {
    regterm_rows <- lapply(block_terms, function(t) {
      rtt <- tryCatch(regTermTest(svy_obj$fit, t, method = "Wald"),
                       error = function(e) NULL)
      if (is.null(rtt)) return(NULL)
      tibble(term = t, svy_F = as.numeric(rtt$Ftest),
             svy_df = rtt$df, svy_ddf = rtt$ddf, svy_p = rtt$p)
    })
    regterm_df <- bind_rows(regterm_rows)
    wald_df <- left_join(wald_df, regterm_df, by = "term")
  }

  wald_df
}
```



# -----------------------------------------------------------------------------
# 11. Build result table + pretty-print (same layout as the prior version)
# -----------------------------------------------------------------------------
```{r}
make_result_df <- function(coef_tbl, sample_label, n, family_used) {
  coef_tbl |>
    mutate(
      z        = coef / se,
      p        = 2 * pnorm(-abs(z)),
      IRR      = exp(coef),
      IRR_lo95 = exp(coef - 1.96 * se),
      IRR_hi95 = exp(coef + 1.96 * se),
      family   = family_used,
      sample   = sample_label,
      outcome  = OUTCOME_COL,
      n        = n
    )
}

print_coef_table <- function(res_df) {
  sections <- list(
    "Race (vs White)"             = "^RACE",
    "Subsidy program"             = "^SUBSIDY",
    "Rent regulation (vs Market)" = "^REGSTATUS",
    "Renter & property controls"  = "^(OLDER_ADULT|SINGLE|CHILDREN|YEARBUILT|UNITS|ROOMS)",
    "Intercept"                   = "^\\(Intercept\\)"
  )

  hdr <- sprintf("  %-32s  %10s  %7s  %6s  %-20s  %s",
                 "term", "coef(log)", "SE", "IRR", "[95% CI]", "sig")

  for (label in names(sections)) {
    pat  <- sections[[label]]
    rows <- res_df |> filter(grepl(pat, term))
    if (nrow(rows) == 0) next

    cat(sprintf("\n  -- %s --\n", label))
    cat(hdr, "\n")

    for (i in seq_len(nrow(rows))) {
      r   <- rows[i, ]
      sig <- dplyr::case_when(
        r$p < 0.001 ~ "***",
        r$p < 0.01  ~ "**",
        r$p < 0.05  ~ "*",
        TRUE        ~ ""
      )
      flag <- if (grepl("AIAN_NHPI|Other_2plus", r$term)) " \u2020" else ""
      ci   <- sprintf("[%.3f, %.3f]", r$IRR_lo95, r$IRR_hi95)
      cat(sprintf("  %-32s  %10.4f  %7.4f  %6.3f  %-20s  %s%s\n",
                  r$term, r$coef, r$se, r$IRR, ci, sig, flag))
    }
  }
}
```



# -----------------------------------------------------------------------------
# 12. Sample sizes (unchanged)
# -----------------------------------------------------------------------------
```{r}
df_sub   <- df |> filter(IS_SUBSIDIZED)
df_unsub <- df |> filter(!IS_SUBSIDIZED)
df_all   <- df

cat(strrep("=", 60), "\n")
cat("ANALYTIC SAMPLE SIZES\n")
cat(strrep("=", 60), "\n")

for (info in list(
  list(label = "Subsidized only",   data = df_sub,   lvls = SUBSIDY_LEVELS_SUB),
  list(label = "Unsubsidized only", data = df_unsub, lvls = SUBSIDY_LEVELS_UNSUB),
  list(label = "All renters",       data = df_all,   lvls = SUBSIDY_LEVELS_ALL)
)) {
  sub <- prep_subset(info$data, info$lvls)
  cat(sprintf("  %-22s  starting n=%s,  complete-case n=%s\n",
              info$label,
              format(nrow(info$data), big.mark = ","),
              format(nrow(sub),       big.mark = ",")))
}
cat("\n")
```



# -----------------------------------------------------------------------------
# 13. Main loop -- fit, extract, test, diagnose, per sample
# -----------------------------------------------------------------------------
```{r}
all_results <- list()
all_svy     <- list()   # svy_obj list (fit + design + terms), for later reuse
pdf(DIAG_PDF, width = 10, height = 8)

samples <- list(
  list(label = "Subsidized",   data = df_sub,   lvls = SUBSIDY_LEVELS_SUB),
  list(label = "Unsubsidized", data = df_unsub, lvls = SUBSIDY_LEVELS_UNSUB),
  list(label = "All",          data = df_all,   lvls = SUBSIDY_LEVELS_ALL)
)

for (s in samples) {
  cat("\n", strrep("=", 60), "\n", sep = "")
  cat(sprintf("%s  --  sample: %s\n", OUTCOME_LABEL, s$label))
  cat(strrep("=", 60), "\n")

  sub <- prep_subset(s$data, s$lvls, standardize = STANDARDIZE)
  n   <- nrow(sub)

  # ---- Missing-outcome diagnostics (before any complete-case filtering) ----
  fml_s <- build_formula(sub)$fml
  cat("\n")
  test_missingness(sub, fml_s)

  # ---- Fit: NB via withReplicates (or fallback), SEs from the SDR design ---
  svy_obj <- fit_survey_model(sub, use_nb)
  cat(sprintf("  n=%s   Model: %s\n", format(n, big.mark = ","), svy_obj$family_used))

  coef_tbl <- get_model_coefs(svy_obj$fit)
  res      <- make_result_df(coef_tbl, s$label, n, svy_obj$family_used)
  print_coef_table(res)

  # ---- IPW sensitivity: does correcting for nonresponse move anything? ----
  if (inherits(svy_obj$fit, "svynb_rep")) {
    cat("\n  -- Sensitivity: inverse-probability-of-response weighting --\n")
    ipw <- tryCatch(ipw_refit(sub, svy_obj$fml),
                    error = function(e) {
                      cat("  (IPW refit failed:", conditionMessage(e), ")\n"); NULL
                    })
    compare_ipw(svy_obj$fit, ipw)
  }

  # ---- Joint significance tests --------------------------------------------
  cat("\n  -- Joint significance per predictor (Wald, design-based vcov) --\n")
  jt <- run_joint_tests(svy_obj)
  if ("svy_F" %in% names(jt)) {
    print(jt |> mutate(across(where(is.numeric), \(x) round(x, 4))), n = Inf)
    cat("  (svy_F/svy_p from regTermTest(); wald_chisq/p is the model-agnostic\n")
    cat("   cross-check -- both should agree on which blocks are significant.)\n")
  } else {
    print(jt |> mutate(across(where(is.numeric), \(x) round(x, 4))), n = Inf)
    cat("  (regTermTest() not attempted -- svymle NB objects are not guaranteed\n")
    cat("   to support it; wald_chisq/p above is the model-agnostic joint test.)\n")
  }

  # ---- DHARMa residual diagnostics (on an equivalent MASS/glm fit) ---------
  # DHARMa needs a model class it knows how to simulate from. svymle NB
  # (svymle-based) and svyglm-with-quasipoisson (no true likelihood to
  # simulate from) are not directly supported. As a diagnostic proxy, fit
  # the same formula with MASS::glm.nb() (if NB) or glm(poisson) (if not),
  # weighted by the primary FW column only (not the full replicate design --
  # this is a shape/fit check, not a source of reported SEs).
  cat("\n  [ DHARMa residual diagnostics (proxy fit: MASS::glm.nb / glm poisson) ]\n")
  diag_fit <- tryCatch({
    if (use_nb) {
      MASS::glm.nb(svy_obj$fml, data = sub, weights = FW)
    } else {
      glm(svy_obj$fml, data = sub, family = poisson(link = "log"), weights = FW)
    }
  }, error = function(e) {
    cat("  (diagnostic fit failed:", conditionMessage(e), ")\n"); NULL
  })

  if (!is.null(diag_fit)) {
    sim_res <- tryCatch(
      suppressWarnings(simulateResiduals(diag_fit, n = 1000, plot = FALSE)),
      error = function(e) {
        cat("  (DHARMa failed:", conditionMessage(e), ")\n"); NULL
      }
    )
    if (!is.null(sim_res)) {
      unif_test <- testUniformity(sim_res,   plot = FALSE)
      disp_test <- testDispersion(sim_res,   plot = FALSE)
      zi_test   <- testZeroInflation(sim_res, plot = FALSE)

      cat(sprintf(
        "  Uniformity  KS p=%.4f  |  Dispersion p=%.4f  |  Zero-inflation p=%.4f\n",
        unif_test$p.value, disp_test$p.value, zi_test$p.value
      ))
      if (disp_test$p.value < 0.05)
        cat("  NOTE: Residual overdispersion remains -- expected to some degree\n",
            "        given quasipoisson's constant-scale approximation, if that\n",
            "        was the fallback used for this sample.\n")
      if (zi_test$p.value < 0.05)
        cat("  NOTE: Zero-inflation test significant -- re-examine before\n",
            "        assuming NB/Poisson is adequate for this sample.\n")

      plot(sim_res, main = sprintf("DHARMa (proxy fit): %s  [%s]",
                                    OUTCOME_COL, s$label))
    }
  }

  all_results[[s$label]] <- res
  all_svy[[s$label]]     <- svy_obj
}

dev.off()
cat(sprintf("\nDHARMa diagnostic plots saved to %s\n", DIAG_PDF))

cat("\n\u2020 AIAN_NHPI and Other_2plus may have n<10 in the subsidized sample; ",
    "interpret coefficients with caution.\n\n")
```



# -----------------------------------------------------------------------------
# 14. Save
# -----------------------------------------------------------------------------
```{r}
results <- bind_rows(all_results)
write_csv(results, OUTPUT_CSV)
cat(sprintf("Saved full coefficient table to %s\n", OUTPUT_CSV))
cat("Done.\n")
```
